---
title: "Primary Care Network"
author: "Cameron Appel"
date: "2024-10-20"
format: gfm
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE)

library(dplyr)
library(magrittr)
```

# Primary Care Network
Primary Care Networks were created in July 2019 to provide accessible and integrated primary, mental health, and community care for patients. The bulk of the PCN workforce consists of Direct Patient Care staff, funded by the Additional Roles Reimbursement Scheme (ARRS), and each PCN has the flexibility and autonomy to determine which roles are required to meet the specific needs of their local populations. Initially, recruitment focused on clinical pharmacists and social prescribing link workers, with more roles being included over subsequent years.

[NHS Digital](https://digital.nhs.uk/data-and-information/publications/statistical/primary-care-network-workforce) provides public CSVs on the PCN workforce at the PCN-level.

We take the average IMD of each practice in the PCN to assign the PCN to a deprivation quintile.


```{r 1}
# List all relevant files from the raw directory
files <- list.files("raw", pattern = "Primary Care Networks.*Individual Level.csv", full.names = TRUE)

# Load and combine all files
pcn <- lapply(files, function(file) {
  df <- read.csv(file)
  df <- df %>% 
    select(CENSUS_YEAR, CENSUS_MONTH,
           PCN_CODE, PCN_NAME, 
           ICB_CODE, ICB_NAME, STAFF_GROUP, 
           FTE)
  return(df)
}) %>% 
  bind_rows()

pcn %<>% group_by(CENSUS_YEAR, PCN_CODE, PCN_NAME, ICB_NAME, STAFF_GROUP) %>% summarise(FTE = sum(FTE, na.rm = TRUE)) %>% rename(Year = CENSUS_YEAR)

pcn$ICB_NAME <- gsub("NHS ", "", pcn$ICB_NAME)
pcn$ICB_NAME <- gsub(" Integrated Care Board", "", pcn$ICB_NAME)
pcn$ICB_NAME <- gsub(" ICB", "", pcn$ICB_NAME)
pcn$ICB_NAME <- gsub("Cornwall and The Isles Of Scilly", "Cornwall and the Isles of Scilly", pcn$ICB_NAME)
pcn$ICB_NAME <- gsub("Hampshire and Isle Of Wight", "Hampshire and Isle of Wight", pcn$ICB_NAME)

pcn %>% head()
```

# Get IMD from fingertipsR
```{r 2}
# Load necessary library
library(fingertipsR)

# Get the profile ID for the Public Health Outcomes Framework
profiles_data <- profiles()
phof_profile <- profiles_data[profiles_data$ProfileName == "Public Health Outcomes Framework", ]
profile_id <- phof_profile$ProfileID[1]

# Get indicators for the Public Health Outcomes Framework profile
indicators_data <- indicators(ProfileID = profile_id)

# Find the IndicatorID for "Deprivation score (IMD 2019)"
indicator_id <- indicators_data$IndicatorID[indicators_data$IndicatorName == "Deprivation score (IMD 2019)"]

# Get the data for the "Deprivation score (IMD 2019)" indicator
IMD <- fingertips_data(IndicatorID = indicator_id, AreaTypeID = 7) %>%
  select(c("ParentCode","Value")) %>% group_by(ParentCode) %>% summarise(IMD = mean(Value))

IMD %<>% rename(PCN_CODE = ParentCode)
# Display the first few rows of the data
head(IMD)
```

```{r 3}
pcn <- merge(pcn, IMD, by = "PCN_CODE") %>%
  group_by(Year) %>%
  mutate(IMD_quintile = ntile(IMD, 5))

# View the updated data frame
head(pcn)

write.csv(pcn, "pcn_workforce.csv", row.names = FALSE)
```