---
title: "Primary Care Equity Datapack"
subtitle: "Data from latest period for {{ params.ICB_NAME }}"
#author: "Health Equity Evidence Centre"
format:
  pptx:
    reference-doc: "template.pptx"
    fig-width: 12
    fig-height: 8
    fig-fullwidth: TRUE
params:
  ICB_NAME: England
---

```{r setup, include=FALSE}

# Setup code
library(ggplot2)
library(showtext)
library(patchwork)
library(ggtext)
font_add_google("Poppins", family = "Poppins")

showtext_auto()

ggplot + theme_set(theme_minimal() + theme(
  axis.title = element_text(size = 80, family = "Poppins"),
  axis.text = element_text(size = 80, family = "Poppins"),
  plot.caption = element_text(size = 60, family = "Poppins"),
  plot.title = element_text(size = 100, family = "Poppins"),
  plot.subtitle = element_text(size = 90, family = "Poppins"),
  panel.grid = element_line(size = 5),
  legend.title = element_text(size = 60, family = "Poppins"),
  legend.text = element_text(size = 60, family = "Poppins"),
  legend.key.width = unit(3.5, "cm"),
  plot.title.position = "plot"
))

update_geom_defaults("point", list(size = 50))

pks <- c(
  "fingertipsR", "tidyverse", "purrr", "tibble",
  "spatstat", "lubridate", "readODS", "janitor", "patchwork", "gdata"
)

sapply(pks, library, character.only = T)
"%ni%" <- Negate("%in%")

df <- read.csv("final_data.csv")
```

## Introduction
Primary care (in contrast to specialty care) is associated with a more equitable distribution of health in populations.

A key role of commissioners is to ensure the equitable distribution of resources across the system.

We present the latest NHS primary care data, using Index of Multiple Deprivation (IMD) to examine inequalities in patient health determinants and outcomes, according to the following categories:

Resources (supply): Payments, Workforce
Population (demand): Disease prevalence, Health-related behaviours
Service quality: QOF achievement
Access: Patient experience, Appointments
Impact on secondary care: Emergency admissions, A&E attendances
For further information or to discuss the results, please contact Dr John Ford

## Inequality in Life Expectancy
Average life expectancy for men is in the least deprived 20%.
      
```{r Life_Expectancy}
year <- 2019

df_filter <- df %>%
  filter(ICB.NAME == params$ICB_NAME) %>%
  filter(Year == year) %>%
  filter(Indicator %in% c("Life_Expectancy_Female", "Life_Expectancy_Male"))

df_filter %>%
  ggplot(., aes(x = Indicator)) +
  geom_linerange(aes(x = Indicator, ymin = quin_5, ymax = quin_1)) +
  geom_point(aes(y = quin_1, color = "Least deprived")) +
  geom_point(aes(y = quin_5, color = "Most deprived")) +
  geom_point(aes(y = avg, color = "National average"), size = 15, shape = 18) +
  scale_x_discrete(labels = c(
    "Life_Expectancy_Female" = "Female",
    "Life_Expectancy_Male" = "Male"
  )) +
  scale_color_manual(values = c("Least deprived" = "#28B788", "Most deprived" = "#007AA8", "National average" = "#A80026")) +
  labs(
    x = NULL, y = NULL,
    title = paste0("Life expectancy by practice, birth cohort 2015-19 (", params$ICB_NAME, ")"),
    subtitle = "Practices in most and least deprived IMD quintiles. Larger circles represent stronger socio-economic gradient.",
    colour = "",
    caption = "Source: NHS England GP Workforce statistics, 2024. © Health Equity Evidence Centre, 2024."
  ) +
  coord_flip() +
  guides(colour = guide_legend(override.aes = list(size = 50, shape = c(16, 16, 18))))
```


