---
title: "Primary Care Equity Datapack"
author: "Health Equity Evidence Centre"
subtitle: "Data from latest period for `r params$ICB_NAME`"
format: 
  markdown:
    code-fold: true
    toc: true
params:
  ICB_NAME: "England"
---

```{r setup}
# Load necessary libraries
library(ggplot2)
library(showtext)
library(patchwork)
library(ggtext)
library(fingertipsR)
library(tidyverse)
library(purrr)
library(tibble)
library(spatstat)
library(lubridate)
library(readODS)
library(janitor)
library(gdata)

# Set up Google font and automatic display for text rendering
font_add_google("Poppins", family = "Poppins")
showtext_auto()

# Set up the ggplot theme
theme_set(
  theme_minimal() +
    theme(
      axis.title = element_text(size = 80, family = "Poppins"),
      axis.text = element_text(size = 80, family = "Poppins"),
      plot.caption = element_text(size = 60, family = "Poppins"),
      plot.title = element_text(size = 100, family = "Poppins"),
      plot.subtitle = element_text(size = 90, family = "Poppins"),
      panel.grid = element_line(size = 5),
      legend.title = element_text(size = 60, family = "Poppins"),
      legend.text = element_text(size = 60, family = "Poppins"),
      legend.key.width = unit(3.5, "cm"),
      plot.title.position = "plot"
    )
)

# Update default settings for geom_point size
update_geom_defaults("point", list(size = 50))

# Import data
df <- read.csv("final_data.csv")

# Define helper functions or custom operators if needed
"%ni%" <- Negate("%in%")

# Optional: suppress warnings and set figure dimensions globally for knitr
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  fig.width = 90,
  fig.height = 25,
  fig.fullwidth = TRUE
)
```

## Introduction

Primary care, in contrast to specialty care, is associated with a more equitable distribution of health across populations. One of the key roles of commissioners is to ensure that resources are distributed equitably across the healthcare system.

This report presents the latest NHS primary care data, using the Index of Multiple Deprivation (IMD) to examine inequalities in patient health determinants and outcomes. We analyze the data according to the following categories:

- **Resources (supply)**: Payments, Workforce
- **Population (demand)**: Disease prevalence, Health-related behaviors
- **Service quality**: Quality and Outcomes Framework (QOF) achievement
- **Access**: Patient experience, Appointments
- **Impact on secondary care**: Emergency admissions, A&E attendances

For further information or to discuss the results, please contact [Dr. John Ford](mailto:j.a.ford@qmul.ac.uk).

## ICB Overview

```{r overview, warning = FALSE}
payments <- read.csv("../data/payments/payments.csv")

england <- payments %>%
  filter(Year == 2019) %>%
  group_by(Year) %>%
  summarise(total = n(), .groups = "drop")

# Calculate total count by Year, ICB.NAME
total_counts <- payments %>%
  filter(Year == 2019) %>%
  group_by(Year, ICB.NAME) %>%
  summarise(total = n(), .groups = "drop")

# Calculate count and percentage by IMD_quintile for all ICBs
result <- payments %>%
  filter(Year == 2019) %>%
  group_by(Year, IMD_quintile, ICB.NAME) %>%
  summarise(n = n(), .groups = "drop") %>%
  left_join(total_counts, by = c("Year", "ICB.NAME")) %>%
  mutate(perc = (n / total) * 100) %>%
  select(Year, ICB.NAME, IMD_quintile, n, perc)

# Filter and organize data for specific ICBs
result <- result %>%
  filter(!is.na(IMD_quintile)) %>%
  filter(ICB.NAME %in% c("Hertfordshire and West Essex", "Bedfordshire, Luton and Milton Keynes", "Norfolk and Waveney", "Mid and South Essex", "Cambridgeshire and Peterborough", "Suffolk and North East Essex")) %>%
  mutate(ICB.NAME = factor(ICB.NAME, levels = c("Mid and South Essex", "Norfolk and Waveney", "Cambridgeshire and Peterborough", "Suffolk and North East Essex", "Bedfordshire, Luton and Milton Keynes", "Hertfordshire and West Essex")))

# Plot: Percentage of practices in each IMD quintile by ICB
result %>%
  ggplot(aes(x = ICB.NAME, y = perc, fill = factor(IMD_quintile))) +
  geom_bar(stat = "identity", position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(
    values = c("5" = "#A80026", "4" = "#D35400", "3" = "#8E44AD", "2" = "#3498DB", "1" = "#00A676"),
    name = "",
    labels = c("Quintile 1 (least deprived)", "2", "3", "4", "Quintile 5 (most deprived)")
  ) +
  labs(
    x = NULL,
    y = NULL,
    title = "Percentage of practices in each IMD quintile by ICB (East of England)"
  ) +
  theme(
    legend.position = "bottom"
  ) +
  coord_flip()
```

Each practice in England is assigned an Index of Multiple Deprivation (IMD) based on the population it serves, divided into deprivation quintiles.

For `r params$ICB_NAME`, approximately **`r round(result[result$ICB.NAME == params$ICB_NAME, ]$perc[5], 0)`%** of practices serve the most deprived quintile of patients in England.

## Inequality in Life Expectancy      
```{r Life_Expectancy}
year <- 2019

df_filter <- df %>%
  filter(ICB.NAME == params$ICB_NAME) %>%
  filter(Year == year) %>%
  filter(Indicator %in% c("Life_Expectancy_Female", "Life_Expectancy_Male"))

df_filter %>%
  ggplot(., aes(x = Indicator)) +
  geom_linerange(aes(x = Indicator, ymin = quin_5, ymax = quin_1)) +
  geom_point(aes(y = quin_1, color = "Least deprived")) +
  geom_point(aes(y = quin_5, color = "Most deprived")) +
  geom_point(aes(y = avg, color = "National average"), size = 15, shape = 18) +
  scale_x_discrete(labels = c(
    "Life_Expectancy_Female" = "Female",
    "Life_Expectancy_Male" = "Male"
  )) +
  scale_color_manual(values = c("Least deprived" = "#28B788", "Most deprived" = "#007AA8", "National average" = "#A80026")) +
  labs(
    x = NULL, y = NULL,
    title = paste0("Life expectancy by practice, birth cohort 2016-20 (", params$ICB_NAME, ")"),
    subtitle = "Practices in most and least deprived IMD quintiles. Larger circles represent stronger socio-economic gradient.",
    colour = "",
    caption = "Source: NHS England GP Workforce statistics, 2024. © Health Equity Evidence Centre, 2024."
  ) +
  coord_flip() +
  guides(colour = guide_legend(override.aes = list(size = 50, shape = c(16, 16, 18))))
```

Average life expectancy for men is in the least deprived 20%.
