---
title: "Practice Closures Trigger Analysis"
author: "Cameron Appel"
date: "2024-08-06"
output: 
  md_document:
    variant: markdown_github
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

# Practice Closures Trigger Analysis

```{r}
library(magrittr)
### Satisfaction analysis

### Workforce analysis
wf15 <- read.csv("../../data/workforce/raw/15_9.csv")
wf23 <- read.csv("../../data/workforce/raw/23_9.csv")

pay15 <- read.csv("../../data/payments/raw/14-15.csv")
pay23 <- read.csv("../../data/payments/raw/22-23.csv")

fte_15 <- wf15$TOTAL_GP_EXTGL_FTE %>%
  as.numeric() %>%
  sum(na.rm = TRUE)

fte_23 <- wf23$TOTAL_GP_EXTGL_FTE %>%
  as.numeric() %>%
  sum(na.rm = TRUE)
```

General practice is in crisis: patient satisfaction is at a record low; the number of fully qualified GPs has dropped from `r fte_15` in 2015  to `r fte_23` in 2023 (full-time equivalent, excluding GPs in training and locums).

```{r}
n_prac_15 <- wf15$PRAC_CODE %>%
  unique() %>%
  length()
n_prac_23 <- wf23$PRAC_CODE %>%
  unique() %>%
  length()

n_prac_15_payments <- pay15$Practice.Code %>%
  unique() %>%
  length()

n_prac_23_payments <- pay23$Practice.Code %>%
  unique() %>%
  length()
```
Over the past decade there has been a move towards providing general practice at scale to benefit from economies of scale and to ensure practices are more sustainable. This has led to a dramatic reduction in the number of practices from `r n_prac_15` in 2015 to `r n_prac_23` in 2023, according to the workforce data.

There are some discrepancies with other data sources, however; according to the payments data, the number of practices has dropped from `r n_prac_15_payments` in 2015 to `r n_prac_23_payments` in 2023.

The reasons for and consequences of practice closures are multifaceted. Practices may merge with other practices for purely organisational reasons or because one practice is failing. Mergers may be purely administrative with limited impact on patients, or they may involve closure of premises or redesign of services.

Practices may also close altogether, leaving thousands of patients without access to primary care until they find another surgery. Practice closures are costly in that they incur interruptions in services for patients. One reason for closure is when the partnership model fails within a practice; this partnership model is increasingly under strain because younger doctors are less keen to become a partner due to the financial and workload commitment.

ICBs and NHS England are often unaware of a failing practice until it is too late to intervene. The ability to identify practices or areas of the country which are at risk of failing general practices would help commissioners to intervene where necessary or plan for patient care in the event of a practice closure.

# Closed practice count
The NHS Payments data contains information on practice closures. We can use this data to identify practices that have closed in the past decade.

Where multiple closure dates are available for a practice, we will keep the earliest date.
```{r}
library(dplyr)

agg <- data.frame()
closure <- data.frame()

for (file in list.files("../../data/payments/raw/")) {
  df <- read.csv(paste0("../../data/payments/raw/", file))[c("Practice.Code", "Practice.Open.Date", "Practice.Close.Date")]

  # assign year
  year <- substr(file, 1, nchar(file) - 4)
  year <- substr(year, 4, nchar(year))
  year <- paste0("20", year)
  df$Year <- year

  print(year)

  df$Practice.Open.Date[df$Practice.Open.Date %in% c("-", "")] <- NA
  df$Practice.Close.Date[df$Practice.Close.Date %in% c("-", "", " ")] <- NA

  # if Practice.Close.Date is not NA, add row to closure
  closure <- df %>%
    filter(!is.na(Practice.Close.Date)) %>%
    select(Practice.Code, Practice.Close.Date, Year) %>%
    bind_rows(closure)

  agg <- bind_rows(closure, df)
}

# remove duplicates, keeping lowest value of year
closure <- closure %>%
  group_by(Practice.Code) %>%
  filter(Year == min(Year))
```