---
title: "Practice Closures Trigger Analysis"
author: "Cameron Appel"
date: "2024-08-06"
output: 
  md_document:
    variant: gfm
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

# Practice Closures Trigger Analysis
In this analysis, we attempt to quantify and categorise the number of GPs practices that have closed in England each year from 2015 to 2023.

```{r Practice count}
library(magrittr)

for (file in list.files("../../data/payments/raw/")) {
  n_prac <- read.csv(paste0("../../data/payments/raw/", file))$Practice.Code %>%
    unique() %>%
    length()

  if (file == "14-15.csv") {
    n_prac_15_payments <- n_prac
  } else if (file == "22-23.csv") {
    n_prac_23_payments <- n_prac
  }

  print(paste0(sub(".csv$", "", file), ": ", n_prac))
}
```
Over the past decade, there has been a dramatic reduction in the number of practices from `r n_prac_15_payments` in 2015 to `r n_prac_23_payments` in 2023, according to the number of unique practice codes in the annual [NHS Payments data](https://github.com/camappel/HEEC/tree/main/data/payments).

The reasons for and consequences of practice closures are multifaceted. Practices may merge with other practices for purely organisational reasons or because one practice is failing. Mergers may be purely administrative with limited impact on patients, or they may involve closure of premises or redesign of services.

Practices may also close altogether, leaving thousands of patients without access to primary care until they find another surgery. Practice closures are costly in that they incur interruptions in services for patients. One reason for closure is when the partnership model fails within a practice; this partnership model is increasingly under strain because younger doctors are less keen to become a partner due to the financial and workload commitment.

We begin by categorising types of closures:

* Merged: Horizontal integration of 2 or more practices. The practice code changes to a pre-existing practice code.
  +	Administrative: premises are unchanged, limited impact on workforce and patients
  +	Estate restructure: premises close, staff are let go and patient satisfaction falls

* Closed: The practice code no longer exists
  + Practice no longer receives payments, has no registered payments, the premises close, and staff are let go.

* Open: The practice has had the same practice code since it opened and still exists in the most recent records

# Methods
### NHS Payments data
The NHS Payments data contains information on practice closures, under the column `Practice.Code.Date`. However, no further information is provided on the type of closure.

```{r Practice close date}
library(dplyr)

closure <- data.frame()

for (file in list.files("../../data/payments/raw/")) {
  df <- read.csv(paste0("../../data/payments/raw/", file))[c("Practice.Code", "Practice.Open.Date", "Practice.Close.Date")]

  # assign year
  year <- substr(file, 1, nchar(file) - 4)
  year <- substr(year, 4, nchar(year))
  year <- paste0("20", year)
  df$Year <- year

  # replace empty values with NA
  df$Practice.Open.Date[df$Practice.Open.Date %in% c("-", "")] <- NA
  df$Practice.Close.Date[df$Practice.Close.Date %in% c("-", "", " ")] <- NA

  # if Practice.Close.Date is not NA, add row to closure dataframe
  closure <- df %>%
    filter(!is.na(Practice.Close.Date)) %>%
    select(Practice.Code, Practice.Close.Date, Year) %>%
    bind_rows(closure)
}

# remove duplicates, keeping lowest value of year
closure <- closure %>%
  group_by(Practice.Code) %>%
  filter(Year == min(Year))

# print number of closed practices per year
closure %>%
  group_by(Year) %>%
  summarise(n = n())
```