---
title: "Exploring the Impact of Dispensing Practices on Equity in NHS Payments to General Practices"
author: "Your Name"
date: "2024-08-06"
output: 
  md_document:
    variant: markdown_github
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

# Exploring the impact of dispensing practicing on equity in NHS payments to general practices

General practices serving the most deprived populations receive less funding per weighted patient than those serving the least deprived. Here we show that this inequality is driven by a higher concentration of dispensing practices in more affluent areas.

### What are Dispensing Practices?

Dispensing practices are practices where “at the patient’s request dispensing doctors are allowed to dispense the medicines they prescribe for these patients” (DDA); they are not the same as practices with an on-site pharmacy.

[NHS Digital](https://digital.nhs.uk/data-and-information/publications/statistical/nhs-payments-to-general-practice) provides public CSVs on NHS payments to individual providers of general practice services in England from 2014-15 to 2022-23. We look at the most recent data from 2022-23.

```{r}
# Load necessary libraries
library(magrittr)
library(dplyr)

# Download the NHS payments data for 2022/23
df <- read.csv("https://files.digital.nhs.uk/05/E6ADA0/nhspaymentsgp-22-23-prac-csv.csv")

# Filter dispensing and non-dispensing practices
dispensing <- df %>% filter(Dispensing.Practice == "Yes")
non_dispensing <- df %>% filter(Dispensing.Practice == "No")

# Number of dispensing practices
num_dispensing <- dispensing %>%
  distinct(Practice.Code) %>%
  nrow()

# Number of non-dispensing practices
num_non_dispensing <- non_dispensing %>%
  distinct(Practice.Code) %>%
  nrow()

num_unknown_dispensing <- df[df$Dispensing.Practice == "Unknown", ] %>%
  distinct(Practice.Code) %>%
  nrow()

# Sum of patients in dispensing practices
total_dispensing_patients <- sum(dispensing$Average.Number.of.Registered.Patient, na.rm = TRUE)

# Sum of patients in non-dispensing practices
total_non_dispensing_patients <- sum(non_dispensing$Average.Number.of.Registered.Patient, na.rm = TRUE)
```

In 2023, there were `r num_dispensing` dispensing practices covering `r total_dispensing_patients` patients and `r num_non_dispensing` non-dispensing practices covering `r total_non_dispensing_patients` patients in England (dispensing status unknown for `r num_unknown_dispensing` practices).

[OHID](https://fingertips.phe.org.uk/search/deprivation%20index#page/4/gid/1/pat/159/par/K02000001/ati/15/are/E92000001/iid/93553/age/1/sex/4/cat/-1/ctp/-1/yrr/1/cid/4/tbm/1) provides the [Fingertips API](https://fingertips.phe.org.uk/api), which includes practice-level IMD values.

```{r 2}
# Load necessary libraries
library(httr)
library(readr)

# Define the URL and query parameters
base_url <- "https://fingertipsws.phe.org.uk/api/all_data/csv/by_indicator_id"
query_params <- list(
  v = "/0-c459298b/",
  parent_area_code = "E92000001",
  parent_area_type_id = 167,
  child_area_type_id = 7,
  indicator_ids = 93553
)

# Make the API request
response <- GET(base_url, query = query_params)

# Check if the response is successful
if (http_status(response)$category == "Success") {
  # Write the content to a temporary file
  temp_file <- tempfile(fileext = ".csv")
  writeBin(content(response, "raw"), temp_file)

  # Read the CSV data
  IMD <- read_csv(temp_file)

  # Display the first few rows of the data
  print(head(IMD))
} else {
  cat("Failed to retrieve data. Status code:", status_code(response), "\n")
  cat("Response content:", content(response, "text"), "\n")
}

# select Area Code, Time period, Value columns and rename them to Practice.Code, Year, IMD
IMD <- IMD %>%
  select("Area Code", "Time period", "Value") %>%
  rename(Practice.Code = "Area Code", Year = "Time period", IMD = "Value")
```

As well as the the fingertipsR package, which provides an R interface to the Fingertips API.

```{r echo = FALSE}
# # Load necessary library
# library(fingertipsR)
#
# # Get the profile ID for the Public Health Outcomes Framework
# profiles_data <- profiles()
# phof_profile <- profiles_data[profiles_data$ProfileName == "Public Health Outcomes Framework", ]
# profile_id <- phof_profile$ProfileID[1]
#
# # Print the profile details (optional)
# print(phof_profile)
#
# # Get indicators for the Public Health Outcomes Framework profile
# indicators_data <- indicators(ProfileID = profile_id)
#
# # Find the IndicatorID for "Deprivation score (IMD 2019)"
# indicator_id <- indicators_data$IndicatorID[indicators_data$IndicatorName == "Deprivation score (IMD 2019)"]
#
# # Print the indicator details (optional)
# print(indicators_data[indicators_data$IndicatorName == "Deprivation score (IMD 2019)", ])
#
# # Get the data for the "Deprivation score (IMD 2019)" indicator
# data <- fingertips_data(IndicatorID = indicator_id, AreaTypeID = 7) # AreaTypeID 7 is for CCGs
#
# # Display the first few rows of the data
# head(data)
```

To explore the relationship between dispensing practices and deprivation, we can merge the NHS payments data with the IMD data (from the API, not the package).

```{r}
# Merge the NHS payments data with the IMD data, starting with 2019 data
df_merged <- merge(df, IMD[IMD$Year == max(IMD$Year), ], by = "Practice.Code")

# Assign each practice to a deprivation quintile based on the IMD score
df_merged %<>%
  mutate(IMD_quintile = cut(IMD, breaks = quantile(IMD, probs = seq(0, 1, 0.2), na.rm = TRUE), include.lowest = TRUE, labels = c("1", "2", "3", "4", "5")))

# How many practices in df are missing from df_merged?
num_missing <- df[!df$Practice.Code %in% df_merged$Practice.Code, ]$Practice.Code %>%
  unique() %>%
  length()

# How many dispensing practices are in each quintile?
num_disp_5 <- table(df_merged$IMD_quintile, df_merged$Dispensing.Practice)[5, 2]

# What is the dispensing status of the practices missing from df_merged?
dispensing_status_missing <- df[!df$Practice.Code %in% df_merged$Practice.Code, ] %>%
  group_by(Dispensing.Practice) %>%
  summarise(
    n = n()
  )
```

IMD values were not available for `r num_missing` practices, including all of the practices with unknown dispensing status, leaving `r nrow(df_merged)` practices with known IMD values and dispensing status out of the original `r nrow(df)` practices.

```{r}
# calculate the average payment per patient for the highest and lowest deprivation quintiles for dispensing and non-dispensing practices
agg_dispensing <- df_merged %>%
  filter(
    (IMD_quintile == 1 | IMD_quintile == 5) &
      (Dispensing.Practice == "No" | Dispensing.Practice == "Yes") &
      !is.na(IMD_quintile)
  ) %>%
  group_by(IMD_quintile, Dispensing.Practice) %>%
  summarise(
    total_payments = sum(Total.NHS.Payments.to.General.Practice.including.Covid.and.PCN.payments, na.rm = TRUE),
    weighted_patients = sum(Average.Number.of.Weighted.Patients, na.rm = TRUE),
    .groups = "drop" # to ungroup after summarise
  ) %>%
  mutate(
    payment_per_patient = total_payments / weighted_patients
  )

# calculate the average payment per patient for all practices
additional_rows <- bind_rows(
  data.frame(
    IMD_quintile = "1",
    Dispensing.Practice = "All",
    total_payments = sum(df_merged[df_merged$IMD_quintile == 1, ]$Total.NHS.Payments.to.General.Practice.including.Covid.and.PCN.payments, na.rm = TRUE),
    weighted_patients = sum(df_merged[df_merged$IMD_quintile == 1, ]$Average.Number.of.Weighted.Patients, na.rm = TRUE)
  ) %>%
    mutate(
      payment_per_patient = total_payments / weighted_patients
    ),
  data.frame(
    IMD_quintile = "5",
    Dispensing.Practice = "All",
    total_payments = sum(df_merged[df_merged$IMD_quintile == 5, ]$Total.NHS.Payments.to.General.Practice.including.Covid.and.PCN.payments, na.rm = TRUE),
    weighted_patients = sum(df_merged[df_merged$IMD_quintile == 5, ]$Average.Number.of.Weighted.Patients, na.rm = TRUE)
  ) %>%
    mutate(
      payment_per_patient = total_payments / weighted_patients
    )
)

agg_dispensing <- bind_rows(agg_dispensing, additional_rows)

# Replace the "All" Dispensing.Practice with "All Practices"
agg_dispensing$Dispensing.Practice[agg_dispensing$Dispensing.Practice == "All"] <- "All practices"
agg_dispensing$Dispensing.Practice[agg_dispensing$Dispensing.Practice == "Yes"] <- "Dispensing Practices"
agg_dispensing$Dispensing.Practice[agg_dispensing$Dispensing.Practice == "No"] <- "Non-dispensing Practices"

# Update the factor levels for Dispensing.Practice
agg_dispensing$Dispensing.Practice <- factor(agg_dispensing$Dispensing.Practice,
  levels = c("Dispensing Practices", "Non-dispensing Practices", "All practices")
)

pwp_1 <- agg_dispensing[5, 5]
pwp_5 <- agg_dispensing[6, 5]
pwp_diff <- pwp_1 - pwp_5

pwp_disp_diff <- agg_dispensing[1, 5] - agg_dispensing[3, 5]
pwp_disp_diff2 <- agg_dispensing[2, 5] - agg_dispensing[4, 5]
```

Across all practices, those in the least deprived 20% received £`r pwp_1` per weighted patient in 2023, compared to £`r pwp_5` for those in the most deprived 20% – a gap of £`r pwp_diff`. When we exclude all dispensing practices from this analysis, the gap is only £`r pwp_disp_diff`. The gap in payments to dispensing practices is much larger (£`r pwp_disp_diff2`), but this is partly because there are only `r num_disp_5` dispensing practices covering the most deprived 20%.

```{r}
library(ggplot2)

# Plot
ggplot(agg_dispensing, aes(x = payment_per_patient, y = Dispensing.Practice, color = as.factor(IMD_quintile))) +
  geom_point(size = 5) +
  scale_color_manual(values = c("5" = "#A80026", "1" = "#1B2C57"), labels = c("Least deprived (Q1)", "Most deprived (Q5)")) +
  labs(
    title = "Total Payment per Weighted Patient, 2022/23",
    subtitle = "Practices in Most and Least Deprived IMD Quintiles",
    x = "Payment per Patient (£)",
    y = "Dispensing Practice",
    color = "IMD Quintile"
  ) +
  theme_minimal(base_size = 15) +
  theme(
    plot.title = element_text(size = 18, face = "bold"),
    plot.subtitle = element_text(size = 14),
    axis.title.y = element_blank(),
    axis.text.x = element_text(size = 10),
    legend.position = "bottom",
    legend.title = element_text(size = 10),
    legend.text = element_text(size = 8),
    legend.key.size = unit(0.5, "cm"),
    panel.grid.major = element_line(color = "grey80"),
    panel.grid.minor = element_line(color = "grey90"),
    plot.margin = margin(5, 5, 5, 5) # Adjust margin to make space for the legend
  )
```


```{r}
table(dispensing$Practice.Rurality)

perc_rural <- table(dispensing$Practice.Rurality) / nrow(dispensing) * 100

table(df_merged[df_merged$Practice.Rurality == "Rural", ]$IMD_quintile)

perc_dep <- table(df_merged[df_merged$Practice.Rurality == "Rural", ]$IMD_quintile) / nrow(df_merged[df_merged$Practice.Rurality == "Rural", ]) * 100
```

`r perc_rural[1]`% of all dispensing practices are in rural areas. One explanation for this unequal geographical distribution is that in order for a dispensing practice to dispense medicine, “the patient must live more than 1.6km from a retail pharmacy or the area must be designated as a ‘reserved location'” (DDE). Therefore, dispensing practices are only likely to dispense enough medicine to be sufficiently profitable in rural areas.

Consequently, dispensing practices tend to be in more affluent areas: `r perc_dep[1]`% of dispensing practices serve patients in the least deprived quintile and `r perc_dep[2]`% serve those in the second least deprived quintile, whereas `r perc_dep[5]`% cover the most deprived quintile.

```{r}
# Define the colors vector as provided
colors <- c("#EF7A34", "#00A865", "#007AA8", "#531A5C", "#A80026")

# Custom labels for the legend
custom_labels <- c(
  "1" = "Q1 (least deprived)",
  "2" = "Q2",
  "3" = "Q3",
  "4" = "Q4",
  "5" = "Q5 (most deprived)"
)

dispensing_plot <- ggplot(df_merged[df_merged$Dispensing.Practice == "Yes", ], aes(x = as.factor(IMD_quintile), fill = as.factor(IMD_quintile))) +
  geom_bar() +
  scale_fill_manual(values = colors, labels = custom_labels) +
  labs(
    title = "Dispensing Practices",
    # x = "IMD Quintile",
    y = "Number of Practices",
    fill = "IMD Quintile"
  ) +
  theme_minimal(base_size = 15) +
  theme(
    plot.title = element_text(size = 18, face = "bold"),
    axis.title.x = element_blank(),
    axis.title.y = element_text(size = 14),
    axis.text.x = element_blank(), # Adjust margin to reduce space
    axis.text.y = element_text(size = 12),
    legend.position = "right",
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 10),
    panel.grid.major.x = element_blank(), # Remove major vertical grid lines
    panel.grid.minor.x = element_blank(), # Remove minor vertical grid lines
    # panel.grid.major = element_line(color = "grey80"),
    # panel.grid.minor = element_line(color = "grey90")
  )

non_dispensing_plot <- ggplot(df_merged[df_merged$Dispensing.Practice == "No", ], aes(x = as.factor(IMD_quintile), fill = as.factor(IMD_quintile))) +
  geom_bar() +
  scale_fill_manual(values = colors, labels = custom_labels) +
  labs(
    title = "Non-dispensing Practices",
    y = "Number of practices",
    fill = "IMD Quintile"
  ) +
  theme_minimal(base_size = 15) +
  theme(
    plot.title = element_text(size = 18, face = "bold"),
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    axis.text.x = element_blank(), # Adjust margin to reduce space
    axis.text.y = element_text(size = 12),
    legend.position = "right",
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 10),
    panel.grid.major.x = element_blank(), # Remove major vertical grid lines
    panel.grid.minor.x = element_blank(), # Remove minor vertical grid lines
    # panel.grid.major = element_line(color = "grey80"),
    # panel.grid.minor = element_line(color = "grey90"),
  )

# Combine the two plots using patchwork
library(patchwork)

combined_plot <- dispensing_plot + non_dispensing_plot +
  plot_layout(guides = "collect") & theme(legend.position = "bottom")

print(combined_plot)
```

## How do payments differ between dispensing and non-dispensing practices?
Dispensing and non-dispensing practices receive reimbursement for the drugs they use.

We calculate the proportion of payments that are for prescribing for dispensing and non-dispensing practices.

```{r}
Prescribing <- c("Prescribing.Fee.Payments", "Dispensing.Fee.Payments", "Reimbursement.of.Drugs")

df_merged$Total.Prescribing <- rowSums(df_merged[, Prescribing], na.rm = TRUE)

agg <- df_merged %>%
  group_by(Dispensing.Practice) %>%
  summarise(
    total_prescribing = sum(Total.Prescribing, na.rm = TRUE),
    total_payments = sum(Total.NHS.Payments.to.General.Practice.including.Covid.and.PCN.payments, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    prescribing_proportion = (total_prescribing / total_payments) * 100
  )

agg
```

For dispensing practices, prescribing payments constitute `r agg[2,4]`% of total payments for dispensing practices, but only `r agg[1,4]` for non-dispensing practices.

```{r}
agg <- df_merged %>%
  group_by(IMD_quintile) %>%
  summarise(
    total_prescribing = sum(Total.Prescribing, na.rm = TRUE),
    total_payments = sum(Total.NHS.Payments.to.General.Practice.including.Covid.and.PCN.payments, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    prescribing_proportion = (total_prescribing / total_payments) * 100
  )

diff_all <- agg[1, 3] - agg[5, 3]
ratio_all <- agg[1, 3] / agg[5, 3]

diff_wo_pres <- (agg[1, 3] - agg[1, 2]) - (agg[5, 3] - agg[5, 2])
ratio_wo_pres <- (agg[1, 3] - agg[1, 2]) / (agg[5, 3] - agg[5, 2])

agg
```
As result, prescribing fees constituted a larger percentage of total payments for more affluent practices (`r agg[1,4]`%) than less affluent ones (`r agg[5,4]`%) in 2023.

Excluding prescribing payments from the total reduces the difference in total payments between practices serving the most and least deprived populations from `r diff_all` to `r diff_wo_pres` (or from a ratio of `r ratio_all` to `r ratio_wo_pres`)

# Do prescribing contribute more to funding inequality than other payment types?
First, we categorise all payments into 7 types: Global Sum, IT & Premises, PCO, QOF, Contracted services, Prescribing, and COVID. We then calculate the proportion of each payment type for IMD quintile, and the correlation of these proportions with the IMD.
```{r}
df_merged %>% colnames()

globalSum <- c(
  "Global.Sum", "MPIG.Correction.factor", "Balance.of.PMS.Expenditure"
)

QOF <- c("Total.QOF.Payments")

contractedServices <- c(
  "Childhood.Vaccination.and.Immunisation.Scheme",
  "GP.Extended.Hours.Access",
  "Influenza.and.Pneumococcal.Immunisations",
  "Learning.Disabilities",
  "Meningitis",
  "Minor.Surgery",
  "Out.Of.Area.in.Hours.Urgent.Care",
  "Pertussis",
  "Rotavirus.and.Shingles.Immunisation",
  "Services.for.Violent.Patients",
  "Medical.Assessment.Reviews",
  "Medical.Assessment.Reviews",
  "Weight.Management.Service",
  "Non.DES.Item.Pneumococcal.Vaccine..Childhood.Immunisation.Main.Programme"
)

ITPremises <- c(
  "Premises.Payments", "Information.Management.and.Technology"
)

PCO <- c(
  "Seniority",
  "Doctors.Retainer.Scheme.Payments",
  "Total.Locum.Allowances",
  "Appraisal...Appraiser.Costs.in.Respect.of.Locums",
  "Prolonged.Study.Leave",
  "PCO.Admin.Other",
  "General.Practice.Transformation",
  "PCN.Participation",
  "winter.Access.Fund",
  "Other.Payments",
  "PCN.Leadership",
  "PCN.Support",
  "PCN.Extended.Hours.Access",
  "PCN.Workforce",
  "PCN.Investment.and.impact.Fund",
  "PCN.Care.Home.Premium",
  "PCN.Enhanced.Access"
)

COVID <- c(
  "Covid.Immunisation", "Covid.Support.and.Expansion",
  "Long.Covid"
)

# setdiff(names(df_merged), c(globalSum, ITPremises, PCO, QOF, contractedServices, Prescribing, COVID))

df_merged$Total.globalSum <- rowSums(df_merged[, globalSum], na.rm = TRUE)
df_merged$Total.ITPremises <- rowSums(df_merged[, ITPremises], na.rm = TRUE)
df_merged$Total.PCO <- rowSums(df_merged[, PCO], na.rm = TRUE)
df_merged$Total.contractedServices <- rowSums(df_merged[, contractedServices], na.rm = TRUE)
df_merged$Total.COVID <- rowSums(df_merged[, COVID], na.rm = TRUE)

agg <- df_merged %>%
  group_by(IMD_quintile) %>%
  summarise(
    total_prescribing = sum(Total.Prescribing, na.rm = TRUE),
    total_globalSum = sum(Total.globalSum, na.rm = TRUE),
    total_ITPremises = sum(Total.ITPremises, na.rm = TRUE),
    total_PCO = sum(Total.PCO, na.rm = TRUE),
    total_QOF = sum(Total.QOF.Payments, na.rm = TRUE),
    total_contractedServices = sum(Total.contractedServices, na.rm = TRUE),
    total_COVID = sum(Total.COVID, na.rm = TRUE),
    total_payments = sum(Total.NHS.Payments.to.General.Practice.including.Covid.and.PCN.payments, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    prescribing_proportion = (total_prescribing / total_payments) * 100,
    globalSum_proportion = (total_globalSum / total_payments) * 100,
    ITPremises_proportion = (total_ITPremises / total_payments) * 100,
    PCO_proportion = (total_PCO / total_payments) * 100,
    QOF_proportion = (total_QOF / total_payments) * 100,
    contractedServices_proportion = (total_contractedServices / total_payments) * 100,
    COVID_proportion = (total_COVID / total_payments) * 100
  )
agg[, c(1, 10:16)]

# calculate the difference between the total payments for the most and least deprived quintiles for each payment type
diffs_abs <- sapply(agg[, c(2:9)], function(x) x[1] - x[5]) %>% sort()
diffs_abs

# calculate the difference between the proportion of payments for the most and least deprived quintiles for each payment type
diffs_prop <- sapply(agg[, c(10:16)], function(x) x[1] - x[5]) %>% sort()
diffs_prop

# calculate the association between IMD_quintile and the proportion of payments for each payment type
correlations <- sapply(df_merged[, c(25, 79:84)], function(x) cor(as.numeric(df_merged$IMD), x)) %>% sort()
correlations
```

The largest difference in payment proportions between the most and least deprived quintiles is for prescribing payments. Furthermore, while all payment types are negatively correlated with IMD, prescribing payments have the highest absolute correlation score with IMD. This suggests that prescribing payments contribute most to the observed inequality in payments between practices serving the most and least deprived populations.

```{r}
library(reshape2)

agg_melted <- agg[, "IMD_quintile"]
agg_melted[, 2] <- agg["total_globalSum"]
agg_melted[, 3] <- agg["total_ITPremises"]
agg_melted[, 4] <- agg["total_contractedServices"]
agg_melted[, 5] <- agg["total_PCO"]
agg_melted[, 6] <- agg["total_QOF"]
agg_melted[, 7] <- agg["total_prescribing"]
agg_melted[, 8] <- agg["total_COVID"]

# divide all values by 1,000,000
agg_melted[, 2:8] <- agg_melted[, 2:8] / 1000000

agg_melted <- agg_melted %>%
  melt(id.vars = "IMD_quintile")
```

```{r}
colors <- c("#A80026", "#531A5C", "#1B2C57", "#007AA8", "#00A865", "#EF7A34", "#FFD700")

ggplot(agg_melted[!is.na(agg_melted$IMD_quintile), ], aes(x = IMD_quintile, y = value, fill = variable)) +
  geom_bar(stat = "identity") +
  labs(
    title = paste0("Percentage of total payments by type by IMD quintile (2023)"),
    x = "Column",
    y = "Total payments (£, millions)"
  ) +
  scale_fill_manual(values = colors, labels = c("Global sum", "IT premises", "Contracted services", "PCO", "QOF", "Prescribing", "COVID")) +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(), # Remove minor vertical grid lines
    panel.grid.major.y = element_blank(), # Remove minor horizontal grid lines
    panel.grid.minor.x = element_blank(), # Remove minor vertical grid lines
    panel.grid.minor.y = element_blank(), # Remove minor horizontal grid lines
    axis.text.x = element_text(vjust = 10) # Rotate x-axis labels and adjust horizontal justification
  ) +
  xlab("IMD quintile") +
  guides(
    fill = guide_legend(title = "Type")
  ) +
  theme(
    axis.title.x = element_text(vjust = 5) # Move x-axis label upwards
  )
```

## References
NHS Digital: NHS Payments to General Practice
Fingertips API Documentation